Redundant calibration as a complex optimization problem

### 2017-11-12 paper.tex
added the almost final version of the paper

\newcommand{\aaps}{A\&AS}
\newcommand{\aap}{A\&A}
\newcommand{\mnras}{MNRAS}
\newcommand{\nat}{Nature}
\newcommand{\physrep}{Phys. Rep.}

@@

\title[Unknown]{Redundant calibration as a complex optimization problem}
\author[T.L.~Grobler et al.]{T.L.~Grobler$^{1}$\thanks{E-mail: email@address (AVR)}, G.~Bernardi$^{1,2}$, Z.S.~Ali$^{3}$?, C.L.~Carilli$^{4,5}$?, J.S.~Dillon$^{3}$?, A.~Liu$^{3}$?, \newauthor
A.R.~Parsons$^{3,6}$? and O.M.~Smirnov$^{1,2}$?\\
$^{1}$Department of Physics and Electronics, Rhodes University, PO Box 94, Grahamstown, 6140, South Africa\\
$^{2}$SKA SA, 3rd Floor, The Park, Park Road, Pinelands, 7405, South Africa\\
$^{3}$Dept. of Astronomy, University of California, Berkeley, CA 94720, USA\\
$^{4}$National Radio Astronomy Obs., Socorro NM\\
$^{5}$Cavendish Lab., Cambridge UK\\
$^{6}$Radio Astronomy Lab., U. California, Berkeley CA 94720, USA}
\begin{document}

@@

\begin{abstract}
%In this paper we show that instead of splitting our redundant calibration problem into its phase-and-amplitude components (the more traditional approach) we can treat the complex unknown model parameters and their complex conjugates as separate unknowns. We also derive the Levenberg-Maquardt and Gauss-Newton update steps associated with this alternative parameter-and-conjugate framework. We focus more on the Levenberg-Maquardt algorithm in this paper as it exhibits better convergence properties when compared to Gauss-Newton. The Gauss-Newton algorithm is currently the de facto standard that is being used to perform redundant calibration. It is common practice to implement redundant calibration via a hierarchical algorithmic strategy. Using the Levenberg-Maquardt algorithm, therefore, reduces the number of algorithmic layers that are needed to perform redundant calibration. We also compare two previously proposed methods to accelerate redundant calibration, namely the alternating direction implicit method and the preconditioned conjugate gradient method. We found that the alternating direction implicit method outperforms the preconditioned conjugate gradient method. However, the preconditioned conjugate gradient method is more robust as it is more straightforward to adapt to other calibration use cases. We make use of the parameter-and-conjugate framework in this paper, rather than the more traditional amplitude-and-phase approach as it makes it trivial to relate the alternating direction implicit method and the Levenberg-Marquardt algorithm with one another.
Observations of the redshifted 21-cm line from the epoch of reionization have recently motivated the construction of low frequency radio arrays with highly redundant configurations that provide an alternative calibration strategy - ''redundant calibration" - and boosts sensitivity on specific scales. We show that redundant calibration can be formulated in the domain of complex functions of a complex variable. This framework reduces the 
\textbf{[GIANNI TO REWRITE AND FIX ABSTRACT]}
has driven advancements in the calibration of radio interferometers and 
Calibration of the new generation of low frequency radio interferometers faces new challenges due to their very wide field of views 
The construction of new radio interferometers, particularly operating at low frequencies and 
In this paper we show that instead of splitting our redundant calibration problem into its phase-and-amplitude components (the more traditional approach) we can treat the complex unknown model parameters and their complex conjugates as separate unknowns. We also derive the Levenberg-Maquardt and Gauss-Newton update steps associated with this alternative parameter-and-conjugate framework. We focus more on the Levenberg-Maquardt algorithm in this paper as it exhibits better convergence properties when compared to Gauss-Newton. The Gauss-Newton algorithm is currently the de facto standard that is being used to perform redundant calibration. It is common practice to implement redundant calibration via a hierarchical algorithmic strategy. Using the Levenberg-Maquardt algorithm, therefore, reduces the number of algorithmic layers that are needed to perform redundant calibration. We also compare two previously proposed methods to accelerate redundant calibration, namely the alternating direction implicit method and the preconditioned conjugate gradient method. We found that the alternating direction implicit method outperforms the preconditioned conjugate gradient method. However, the preconditioned conjugate gradient method is more robust as it is more straightforward to adapt to other calibration use cases. We make use of the parameter-and-conjugate framework in this paper, rather than the more traditional amplitude-and-phase approach as it makes it trivial to relate the alternating direction implicit method and the Levenberg-Marquardt algorithm with one another.
\textbf{[Could we add JS Kenyon and SJ Wijnholds to author list?]}
\end{abstract}

%that \textsc{lincal} is effectively the non-linear least-squares gradient-based Gauss-Newton algorithm. Therefore, an easy way to reduce the number of layers used in performing redundant calibration is to employ the Levenberg-Marquardt algorithm instead (which would improve the convergence properties of stock standard \textsc{lincal})

\begin{keywords}
circumstellar matter -- infrared: stars.
\end{keywords}

\section{Introduction}
The effects (i.e. the ionosphere, antenna gains, etc.) a celestial signal experiences along its propagation path are modelled with the radio interferometric measurement equation (RIME) \citep{ME1,RRIME1}.
Calibration is the procedure by which we remove the errors that these propagation effects introduce into interferometric data and is normally accomplished using a 
least-squares solver. In the case where we only model the antenna gains, which is the most basic form of the RIME, we can define calibration as finding the antenna gains that minimize the difference between our observed and predicted visibilities.

There are two recent developments in skymodel-based calibration which should be explicitly mentioned at this point, since they form the foundation this paper builds on. The first is the discovery of \textsc{StEfCal} (Statistically Efficient Calibration) \citep{Mitchell:MWA-cal,Salvini2014}, which 
is an alternating direction implicit (ADI) skymodel-based calibration algorithm. It works by assuming the gains are known while solving for their conjugates and then performing the reverse. 
This procedure results in the linearization of the calibration algorithm and reduces the computational complexity of calibration from $\mathbb{O}(N^3)$ to $\mathbb{O}(N^2)$, where $N$ denotes the number of antennas in the array. 
The second development is the idea: instead of breaking the calibration problem 
into its real and imaginary parts and then solving for the real and imaginary parts of the unknown parameters separately we can treat the antenna gains and their 
conjugates as separate parameters to estimate \citep{Smirnov2015}. \citet{Smirnov2015} also showed that the \textsc{StEfCal} algorithm itself is merely a special case of this
gain-and-conjugate framework.

Two baselines are redundant if their baseline difference vectors have the same length and orientation, which imply that they measure the same $uv$-modes. When we use a redundant array layout, we dramatically 
reduce the number of unknowns. If our array is redundant enough we reduce the number of unknowns to such an extent that we are able to solve for the unknown 
observed visibilities themselves in addition to the antenna gains. Redundant calibration is the procedure by which we estimate both the unknown observed visibilities and the antenna gains.

Redundant calibration was popularized in the 1980's. \citet{Noordam1982} used redundant calibration and the WSRT (Westerbork Synthesis Radio Telescope) to produce high-dynamic range images of 3C84.
The first implementations of redundant calibration made use of the logarithm function, because when the logarithm is applied to the scalar measurement equation, redundant 
calibration is transformed into a linear problem \citep{Wieringa1992,Camps2003,Liu2010}. The implementation associated with this approach is commonly referred to as \textsc{logcal}.
We briefly discuss \textsc{logcal} in Appendix~\ref{sec:logcal}. The next improvement in redundant calibration was to perform a first order Taylor expansion, which is equivalent to the Gauss-Newton (GN) algorithm \citep{Kurien2016}, of the measurement equation and then to solve for the
perturbations of the model-parameters \citep{Liu2010}. \citet{Liu2010} perform the first order Taylor expansion of the measurement equation by splitting the unknown parameters into their phase-and-amplitude
components. The implementation of  this approach is called \textsc{lincal}. \textsc{lincal} is derived from first principals in Appendix~\ref{sec:lincal}. 
Currently, redundant calibration is usually implemented using a layered algorithmic strategy. \textsc{logcal} and \textsc{lincal} are both implemented in \textsc{omnical}\footnote{https://github.com/jeffzhen/omnical}, which is a layered redundant calibration pipeline \citep{Zheng2014,Ali2015}.
The solutions of \textsc{logcal} are used as initial input for \textsc{lincal}, since the GN algorithm requires that its starting parameter guesses be in the vicinity 
of the true solution.

It is also interesting to mention here that a completely different calibration strategy can also be used to calibrate redundant arrays, \textsc{CorrCal} \citep{Sievers2017}\footnote{https://github.com/sievers/corrcal2}. In \textsc{CorrCal} the unknown 
visibilities are no longer solved for. The sky is actually modelled, via two covariance matrices. 
The first covariance matrix is associated with known point sources. The second covariance matrix models the remaining sky in a statistical sense. The unknown gains, however, are still 
estimated. We do not elaborate any further on this approach in this paper.

New instruments like LOFAR (Low Frequency Array) \citep{Noorishad2012}, PAPER (Precision Array for Probing the Epoch of Reionization) \citep{Ali2015} and HERA (Hydrogen Epoch of Reionization Array) \citep{deboer2015} have also sparked a renewed interest in redundant calibration. Using redundancy is especially advantageous when we are trying to detect the EoR (Epoch of Reionization) 21 cm signal,
which is inherently quite faint \citep{Parsons2012}. Since redundant baselines make statistically independent measurements of the same $uv$-modes redundancy can greatly 
improve the SNR (signal-to-noise ratio) of EoR detection experiments. This explains why PAPER and HERA are highly redundant arrays, as they are both primarily used for EoR 21 cm detection experiments.

\noindent
\textbf{[GIANNI TO REWRITE AND EXTEND EOR SECTION]}

In this paper we propose to use the complex framework introduced by \citet{Smirnov2015} instead of the traditional amplitude-and-phase approach used by \textsc{lincal}.
We also propose to use the Levenberg-Marquardt (LM) gradient-based algorithm instead of the Gauss-Newton (GN) algorithm as it exhibits better convergence properties.
Using the LM algorithm reduces the number of algorithmic layers that are needed to perform redundant calibration.

Two ways of speeding up redundant calibration have recently been proposed. The first technique makes use of the ADI method, in which we alternate between solving the gains, their conjugates and 
the true observed visibilities \citep{Wijnholds2012,Marthi2014}. Note that the fact that \citet{Marthi2014} make use of the ADI method is not immediately clear
from the paper itself. We will elaborate on this in Section~\ref{sec:adi}. The second is to use the preconditioned conjugate gradient (PCG) method. PCG allows us to exploit the sparsity inherent in the redundant calibration problem 
to speed it up \citep{Liu2010}. A proper comparison between these two approaches has not been done before.
Therefore, we also compare the ADI \citep{Marthi2014} and PCG \citep{Liu2010} methods in this paper. We decided to focus on the complex framework proposed by \citet{Smirnov2015} rather than the traditional phase-and-amplitude approach as it 
allows us to trivially prove that the ADI method proposed by \citep{Marthi2014} is merely a special case of the more general LM algorithm.

%Moreover, this framework also 
%allows us to show that the ADI method proposed by \citep{Marthi2014} can for all intents and purposes be renamed to redundant \textsc{StEfCal} \citep{Mitchell:MWA-cal,Salvini2014}.

In Section~\ref{sec:sky_wirtinger}, we review the skymodel-based complex-and-conjugate calibration approach proposed by \citet{Smirnov2015}.
In Section~\ref{sec:red_wirtinger}, we show that this same approach can also be used to perform redundant calibration.
In Section~\ref{sec:results}, we compare the computational complexity of two recently proposed redundant calibration speedup algorithms with one another.
We end the paper with some final conclusions.

% Main contributions of the paper:

>>>>

\newcommand{\aaps}{A\&AS}
\newcommand{\aap}{A\&A}
\newcommand{\mnras}{MNRAS}
\newcommand{\nat}{Nature}
\newcommand{\apj}{The Astrophysical Journal}
\newcommand{\prd}{Physical Review Letters D}
\newcommand{\physrep}{Phys. Rep.}
\newcommand{\araa}{Annual Review of Astronomy and Astrophysics}
\newcommand{\pasp}{Publications of the Astronomical Society of Pacific}
\newcommand{\procspie}{Proceedings of the SPIE}
\newcommand{\pasa}{Publications of the Astronomical Society of Australia}

@@

%\title[Complex redundant calibration]{Redundant calibration as a complex optimization problem}
\title[Redundant interferometric calibration as a complex optimization problem]{Redundant calibration as a complex optimization problem}
%\author[T.L.~Grobler et al.]{T.L.~Grobler$^{1}$\thanks{E-mail: email@address (AVR)}, G.~Bernardi$^{1,2}$, Z.S.~Ali$^{3}$?, C.L.~Carilli$^{4,5}$?, J.S.~Dillon$^{3}$?, A.~Liu$^{3}$?, \newauthor
%A.R.~Parsons$^{3,6}$? and O.M.~Smirnov$^{1,2}$?\\
%$^{1}$Department of Physics and Electronics, Rhodes University, PO Box 94, Grahamstown, 6140, South Africa\\
%$^{2}$SKA SA, 3rd Floor, The Park, Park Road, Pinelands, 7405, South Africa\\
%$^{3}$Dept. of Astronomy, University of California, Berkeley, CA 94720, USA\\
%$^{4}$National Radio Astronomy Obs., Socorro NM\\
%$^{5}$Cavendish Lab., Cambridge UK\\
%$^{6}$Radio Astronomy Lab., U. California, Berkeley CA 94720, USA}
\author[T.L.~Grobler et al.]{T.L.~Grobler$^{1,2}$\thanks{E-mail: email@address (AVR)}, G.~Bernardi$^{2,3}$, C.L.~Carilli$^{4,5}$, J.S.~Kenyon$^{2}$, A.R.~Parsons$^{6,7}$? \newauthor and O.M.~Smirnov$^{2,3}$\\
$^{1}$Dept of Mathematical Sciences, Computer Science Division, Stellenbosch University, Private Bag X1, 7602 Matieland, South Africa\\
$^{2}$Department of Physics and Electronics, Rhodes University, PO Box 94, Grahamstown, 6140, South Africa\\
$^{3}$SKA SA, 3rd Floor, The Park, Park Road, Pinelands, 7405, South Africa\\
$^{4}$National Radio Astronomy Obs., Socorro NM\\
$^{5}$Cavendish Lab., Cambridge UK\\
$^{6}$Dept. of Astronomy, University of California, Berkeley, CA 94720, USA\\
$^{7}$Radio Astronomy Lab., U. California, Berkeley CA 94720, USA}
\begin{document}

@@

\begin{abstract}
Observations of the redshifted 21-cm line from the epoch of reionization have recently motivated the construction of low 
frequency radio arrays with highly redundant configurations that provide an alternative calibration strategy - 
''redundant calibration" - and boosts sensitivity on specific spatial scales. In this paper we formulate 
calibration of redundant interferometric arrays as a complex optimization problem \citep[i.e.,][]{Smirnov2015}. 
We then solve this optimization problem via the Levenberg-Marquardt algorithm. This calibration approach is more 
robust to initial conditions than current algorithms and, leveraging on an approximate matrix inversion, allows for further optimization and an efficient implementation (``redundant \textsc{StEfCal}"). We also investigated the use of the preconditioned conjugate gradient method as an alternative optimization strategy, instead of 
an approximate matrix inverse, but found that its computational performance is not competitive with respect to ``redundant \textsc{StEfCal}". The efficient implementation of this new algorithm is made publicly available.
%The construction of new radio interferometers, particularly operating at low frequencies and In this paper we show that instead of splitting our redundant calibration problem into its phase-and-amplitude components (the more traditional approach) we can treat the complex unknown model parameters and their complex conjugates as separate unknowns. We also derive the Levenberg-Maquardt and Gauss-Newton update steps associated with this alternative parameter-and-conjugate framework. We focus more on the Levenberg-Maquardt algorithm in this paper as it exhibits better convergence properties when compared to Gauss-Newton. The Gauss-Newton algorithm is currently the de facto standard that is being used to perform redundant calibration. It is common practice to implement redundant calibration via a hierarchical algorithmic strategy. Using the Levenberg-Maquardt algorithm, therefore, reduces the number of algorithmic layers that are needed to perform redundant calibration. We also compare two previously proposed methods to accelerate redundant calibration, namely the alternating direction implicit method and the preconditioned conjugate gradient method. We found that the alternating direction implicit method outperforms the preconditioned conjugate gradient method. However, the preconditioned conjugate gradient method is more robust as it is more straightforward to adapt to other calibration use cases. We make use of the parameter-and-conjugate framework in this paper, rather than the more traditional amplitude-and-phase approach as it makes it trivial to relate the alternating direction implicit method and the Levenberg-Marquardt algorithm with one another.
\end{abstract}

\begin{keywords}
instrumentation: interferometers -- methods: data analysis -- methods: numerical --techniques: interferometric -- cosmology: observations
\end{keywords}

\section{Introduction}
The quest for the redshifted 21-cm line from the Epoch of Reionization (EoR) is a frontier of modern observational cosmology and has motivated the construction of a series of new interferometric arrays operating at low frequencies over the last decade. EoR measurements are challenging because of the intrinsic faintness of the cosmological signal \citep[see, for instance,][for recent reviews]{Furlanetto2016,McQuinn2016} buried underneath foreground emission which is a few orders of magnitude brighter than the EoR anywhere in the sky \citep[e.g.,][]{Bernardi2009,Bernardi2010,Ghosh2012,Dillon2014,Parsons2014}. 
%and that can be separated by leveraging only on the different spectral properties (citations needed!!!). 
As miscalibrated foreground emission leads to artifacts that can jeopardize the EoR signal \citep[e.g.,][]{grobler2014,barry2016,ewall-wice2016}, an exquisite interferometric calibration is required. Such calibration needs an accurate knowledge of both the sky emission and the instrumental response \citep[e.g.,][]{Smirnov2011c}, unless arrays are deployed in a redundant configuration, i.e. with receiving elements placed on regular grids so that they measure many times the same sky brightness emission. Leveraging on this property, redundancy is a promising path to achieve high accurate interferometric calibration \citep[][]{Noordam1982,Wieringa1992,Pearson1984,Liu2010,Noorishad2012,Marthi2014,Sievers2017}. Moreover, redundant configurations provide a sensitivity boost on a selected sample of spatial scales that can be tuned to be the highest signal--to--noise ratio (SNR) EoR modes \citep[][]{Parsons2012,Dillon2016}. These reasons motivated the design and deploment of EoR arrays in redundant configurations like the MIT-EoR \citep{Zheng2014}, the Precision Array to Probe the Epoch of Reionization \citep[PAPER,][]{Ali2015} and the Hydrogen Epoch of Reionization Array \citep[HERA,][]{deboer2017}.

In this paper we present a new algorithm to calibrate redundant arrays based on the complex optimization formalism recently introduced by \cite{Smirnov2015}. 
With respect to current algorithms, it is more robust to initial conditions, while remaining comparatively fast. We also show that given certain approximations this new algorithm reduces to the redundant calibration equivalent of the \textsc{StEfCal} algorithm \citep{Salvini2014}.
%In this paper we will refer to this simplified version of the new algorithm as redundant StEfCal. 
We investigate the speed-up that the preconditioned conjugate gradient (PCG) method provides if it is employed by the new algorithm \citep{Liu2010}.
A comparison between the computational complexity of the optimized new algorithm and redundant StEfCal is also performed.

The paper is organized as follows: we review the complex calibration formalism by \citet{Smirnov2015} in Section~\ref{sec:sky_wirtinger}, we extend it to the redundant case in Section~\ref{sec:red_wirtinger}, we present some computational complexity results in Section~\ref{sec:pcg} and we conclude in Section~\ref{sec:conclusions}.

%\section{Skymodel-based Wirtinger Calibration}
%\section{Calibration described through the Wirtinger formalism}
\section{Wirtinger calibration}
\label{sec:sky_wirtinger}

In a radio interferometer, the true sky visibilities $y_{pq}$ measured by a baseline formed by antenna $p$ and $q$ are always ``corrupted" by the non-ideal response of the receiver, which is often incorporated into a single, receiver-based complex number $g$. The observed visibility $d_{pq}$ is therefore given by \citep{ME1,ME2,RRIME1}
\begin{equation}
\label{eq:vis_definition}
d_{pq} = g_{p}\conj{g_q} \, y_{pq} + n_{pq},
\end{equation}
where $\conj{(*)}$ indicates conplex conjugation and $n_{pq}$ is the thermal noise component. 
The real and imaginary components of the thermal noise are normally distributed with a mean of zero and a
standard deviation $\sigma$:
\begin{equation}
\sigma \propto \frac{T_{\textrm{sys}}}{\sqrt{\Delta \nu \tau}},
\end{equation}
where $T_{\textrm{sys}}$ is equal to the system temperature, $\Delta \nu$ is the observational bandwidth and $\tau$ is the integration time per visibility.

Considering the number of visibilities $B$ (i.e. baselines) measured by an array of $N$ elements $B = \frac{N^2-N}{2}$, equation~\ref{eq:vis_definition} can be expressed in the following vector form:
\begin{equation}
\label{eq:vis_linear_definition}
\bd = \bv + \bn, 
\end{equation}
where 
\begin{align}
 \left [ \bd \right]_{\alpha_{pq}} &= d_{pq}, & \left [ \bv \right ]_{\alpha_{pq}} &= v_{pq}=g_p y_{pq} \conj{g_q},\nonumber\\
 \left [ \bn \right ]_{\alpha_{pq}} &= n_{pq}, &  &\label{eq:vec_linear_definitions}
\end{align}
and 
\begin{equation}
\alpha_{pq} =
\begin{cases}
(q-p) + (p-1)\left (N-\frac{1}{2}p \right ) & \textrm{if}~p<q\\
0 & \textrm{otherwise}
\end{cases}.
\end{equation}
The function $\alpha_{pq}$ therefore maps composite antenna indexes to unique single indexes, i.e:
\begin{equation}
\{\alpha_{12},\alpha_{13},\cdots,\alpha_{N-1N}\} = \{1,2,\cdots,B\} 
\end{equation}
The vectors in equation~\ref{eq:vec_linear_definitions} are column vectors of size $B$ (i.e. $p<q$).

Radio interferometric calibration aims to determine the best estimate of $\bg = [g_1,g_2,\cdots,g_N]^T$ in order to correct the data and, following equation~\ref{eq:vis_linear_definition}, can be formulated as a non linear least-squares optimization problem:
\begin{equation}
\label{eq:least_squares}
\min_{\bg} \Lambda(\bg) = \min_{\bg} \|\br\|_F^2 = \min_{\bg} \|\bd - \bv(\bg)\|_F^2, 
\end{equation}
where $\Lambda$ is the objective function, $\br$ is the residual vector and $\|\cdot\|_F$ denotes the Frobenius norm. In standard interferometric calibration, $y_{pq}$ is assumed to be known at some level, for instance through the observation of previously known calibration sources. The knowledge of $y_{pq}$ is often named ``sky model".

The gradient-based minimization algorithms that are generally used to solve non-linear least-squares problems are generally solved by using gradient-based minimization algorithms (i.e. Gauss--Newton -- GN -- or Levenberg--Marquardt -- LM) that require the model ($\bv$ in equation~\ref{eq:least_squares}) to be differentiable towards each parameter. 
When the least squares problem is complex, it becomes less straightforward to apply these gradient-based minimization methods, as
many complex functions are not differentiable if the classic notion of differentiation is used, i.e. $\frac{\partial \conj{z}}{\partial z}$ does not exist if $z \in \mathbb{C}$.

In order to circumvent the differentiability conundrum associated with complex least squares problems, standard interferometric calibration divides the complex optimization problem into its real and imaginary parts and solves for the real and imaginary parts of the unknown model parameters separately. \citet{Smirnov2015} showed, however, that this approach is not needed if complex calculus \citep{Wirtinger1927} is directly adopted. The Wirtinger derivatives are defined as:
\begin{align}
\label{eq:wir}
\frac{\partial}{\partial z} &= \frac{1}{2}\left ( \frac{\partial}{\partial x} -  i \frac{\partial}{\partial y} \right ),&\frac{\partial}{\partial \conj{z}} &= \frac{1}{2}\left ( \frac{\partial}{\partial x} +  i \frac{\partial}{\partial y} \right ), 
\end{align}
which lead to the following relations:
\begin{align}
\label{eq:wir_z}
\frac{\partial z}{\partial z} & = 1, & \frac{\partial \conj{z}}{\partial z}&=0, & \frac{\partial z}{\partial \conj{z}} & = 0, & \frac{\partial \conj{z}}{\partial \conj{z}}&=1.
\end{align}
If the gradient operator is defined using equation~\ref{eq:wir}, the model $\bv$ now becomes analytic in both $\bg$ and $\conj{\bg}$ and equation \ref{eq:wir_z} can be used to derive the complex variants of the real-valued GN and LM algorithms. In the complex GN and LM algorithms, complex parameters and their conjugates are treated as separate variables.

Assuming that $y_{pq}$ is known, equation~\ref{eq:least_squares} is recast as \citep{Smirnov2015}:
\begin{equation}
\label{eq:least_squares_augmented}
\min_{\breve{\bg}} \Lambda(\breve{\bg}) = \min_{\breve{\bg}} \|\breve{\br}\|_F^2 = \min_{\breve{\bg}} \|\breve{\bd} - \breve{\bv}(\breve{\bg})\|_F^2, 
\end{equation} 
where $\breve{\br} = [\br^T,\conj{\br}^T]^T$, $\breve{\bd} = [\bd^T,\conj{\bd}^T]^T$, $\breve{\bv} = [\bv^T,\conj{\bv}^T]^T$ and $\breve{\bg} = [\bg^T,\conj{\bg}^T]^T$.

The complex GN update is therefore defined as:
\begin{equation}
\label{eq:GN_update_skymodel}
 \Delta \breve{\bg} = (\bJ^H\bJ)^{-1}\bJ^H\breve{\br},
\end{equation}
with 
\begin{equation}
\label{eq:Jacobian_skymodel}
\bJ = \begin{bmatrix}
       \bmJ & \bmJ^*\\
       \conj{\bmJ}^* & \conj{\bmJ} 
      \end{bmatrix},
\end{equation}
and
\begin{align}
\label{eq:jac_entries}
[\bmJ]_{\alpha_{pq},i} &= \frac{\partial v_{pq}}{\partial g_i}, & [\bmJ^*]_{\alpha_{pq},i} &= \frac{\partial v_{pq}}{\partial \conj{g}_i}. 
\end{align}

The complex LM update is very similar, the major difference being a damping factor $\lambda$ is introduced:
\begin{equation}
\label{eq:LM_update_skymodel}
\Delta \breve{\bg} = (\bJ^H\bJ + \lambda\bD)^{-1}\bJ^H\breve{\br},
\end{equation}
where $\bD=\bI\odot\bJ^H\bJ$. In this paper, the identity matrix is denoted by $\bI$. Moreover, the Hadamard product and the Hermition transpose are denoted by $\odot$ and $(\cdot)^H$. Note the use of the Wirtinger derivatives in equation~\ref{eq:jac_entries}.
We will refer to $\bJ^H\bJ$ as the Hessian matrix $\bH$ and to $\bJ^H\bJ + \lambda\bD$ as the modified Hessian matrix $\bmH$ throughout this paper.

Equation~\ref{eq:GN_update_skymodel} or~\ref{eq:LM_update_skymodel} can now be used iteratively to update the parameter vector $\breve{\bg}$:
\begin{equation}
\label{eq:update_skymodel}
\breve{\bg}_{k+1} = \breve{\bg}_{k} + \Delta \breve{\bg}_{k},
\end{equation}
until convergence is reached.

In the case of the GN algorithm, the parameter update step simplifies and becomes \citep{Smirnov2015}
\begin{equation}
\label{eq:one_half}
\breve{\bg}_{k+1} = (\bJ^H\bJ)^{-1}\bJ^H\breve{\bd} + \frac{1}{2}\breve{\bg}_{k}. 
\end{equation}

\citet{Smirnov2015} realized that the diagonal entries of the Hessian matrix $\bH$ are much more significant than its off-diagonal entries, i.e. $\bH$ is nearly diagonal. By approximating $\bH$ by its diagonal and substituting the approximate Hessian matrix the LM parameter update step becomes \citep{Smirnov2015}:
\begin{align}
\breve{\bg}_{k+1} &\approx \frac{1}{1+\lambda}\widetilde{\bH}^{-1}\bJ^H\breve{\bd} + \frac{\lambda}{1+\lambda} \breve{\bg}_k, \nonumber \\
 &= \alpha \widetilde{\bH}^{-1}\bJ^H\breve{\bd} + (1-\alpha)\breve{\bg}_k, 
\label{eq:stef_alpha}  
\end{align}
where $\alpha = \frac{1}{1+\lambda}$. Note that equation~\ref{eq:one_half} and equation~\ref{eq:stef_alpha} are   
not dependent on $\breve{\br}$.

Interestingly enough, if $\lambda = 0$ we obtain the odd parameter update step of \textsc{StEfCal}\footnote{$k\in\{0,2,\cdots\}$}, and if $\lambda=1$ (which corresponds
to $\alpha=\frac{1}{2}$) we obtain the even parameter update step of \textsc{StEfCal}\footnote{$k\in\{1,3,\cdots\}$} \citep[\textsc{StEfCal},][]{Mitchell:MWA-cal,Salvini2014}. In the \textsc{StEfCal} algorithm, the measurement equation, equation~\ref{eq:vis_definition}, is linearized by assuming that the gains are known, but that their conjugates are not. Under this assumption, the system of equations become linear and the conjugates of the gains can be obtained straightforwardly. Starting from the latest value of the gain conjugates, an updated estimate of the gains can be obtained and the succession continues until convergence is reached.
Alternating between solving and fixing different sets of parameters (which is exactly what \textsc{StEfCal} does) is in general referred to 
as the alternating direction implicit method (ADI). The \textsc{StEfCal} algorithm reduces the computational complexity of calibration from $O(N^3)$ to $O(N^2)$.

\section{Redundant Wirtinger Calibration}
\label{sec:red_wirtinger}
Interferometric baselines are named redundant when they sample the exact same visibilities in the $uv$-plane, i.e. if baseline $pq$ and $rs$ are redundant then $y_{pq} = y_{rs}$. It is convenient to group redundant visibilities together and label each group using a single index rather than using their antenna pairs as in equation~\ref{eq:vis_definition}. We introduce a function $\phi_{pq}$ that maps the antenna pair associated with a specific baseline to the single group index of the redundant baseline group that includes the baseline in question, i.e. if baseline $pq$ and $rs$ are redundant then $\phi_{pq} = \phi_{rs}$. The function $\phi_{pq}$ is not symmetric as $\phi_{pq} = 0$ if $p \geq q$. Equation~\ref{eq:vis_definition} can be re-written for a redundant array as:
\begin{equation}
\label{eq:vis_red}
d_{pq} = g_{p}\conj{g_q}y_{\phi_{pq}} + n_{pq},
\end{equation}
with the same vector form as equation~\ref{eq:vis_linear_definition} if
\begin{align}
 \left [ \bd \right]_{\alpha_{pq}} &= d_{pq}, & \left [ \bv \right ]_{\alpha_{pq}} &= v_{pq}=g_p y_{\phi_{pq}} \conj{g_q},\nonumber\\
 \left [ \bn \right ]_{\alpha_{pq}} &= n_{pq}, &  &\label{eq:vec_definitions}
\end{align}
where the vectors in Equation~\ref{eq:vec_definitions} are column vectors of size $B$ (i.e. $p<q$).

We also introduce the following symmetric variant of $\phi_{pq}$:
\begin{equation}
\zeta_{pq} = 
\begin{cases}
\phi_{pq}~\textrm{if}~p \leq q\\
\phi_{qp}~\textrm{if}~p>q
\end{cases},
\end{equation} 
and we will refer to $\zeta_{pq}$ as the symmetric geometric function.
It is possible to construct a simple analytic expression for $\zeta_{pq}$ for an east-west regular array, i.e. $\zeta_{pq} = |q-p|$. It becomes, however, increasingly difficult to construct analytic expressions of $\zeta_{pq}$ for more complicated array layouts. The empirically constructed symmetric geometry functions of three different redundant layouts are displayed in Figure~\ref{fig:geometry_function}. We denote the range of $\zeta_{pq}$ with $\mathcal{R}(\zeta_{pq})$. The maximal element that $\zeta_{pq}$ can ascertain is denoted by $L$ and can be interpreted as the maximal number of unique redundant baseline groups which can be formed for a given array layout.

\subfigure[Hexagonal: $\zeta_{pq}$]
{\includegraphics[width=0.33\textwidth]{./HEX_phi.pdf}\label{fig:HEX_phi}}
\subfigure[Square: $\zeta_{pq}$]
{\includegraphics[width=0.33\textwidth]{./SQR_phi.pdf}\label{fig:SQR_phi}}
\subfigure[Regular east-west: $\zeta_{pq}$]
{\includegraphics[width=0.33\textwidth]{./REG_phi.pdf}\label{fig:REG_phi}}
\caption{Three different redundant antenna layouts: hexagonal (top left), square (top middle) and regular east-west (top right) with their respective symmetric redundancy geometry functions $\zeta_{pq}$ (bottom panels). We used 91 antennas to construct the hexagonal layout, while 100 antennas were used in the square and east-west layouts. In the case of the east-west layout we only plot the positions of the first ten antennas. The maximal amount of redundant baseline groups $L$ that can be formed for 
the hexagonal, square and east-west layouts are 165, 180 and 99 respectively. The analytic expressions of $L$ for an hexagonal, square and east-west layout is $L = 2N-\frac{1}{2}\sqrt{12N-3}-\frac{1}{2}$,
$L=2N-2\sqrt{N}$ and $L=N-1$.\label{fig:geometry_function}}
\end{figure*}

We can now formulate redundant calibration as a least-squares problem:
\begin{equation}
\label{eq:least_squares_red}
\min_{\bz} \Lambda(\bz) = \min_{\bz} \|\br\|_F^2 = \min_{\bz} \|\bd - \bv(\bz)\|_F^2, 
\end{equation}
where
\begin{align}
 \bg &=[g_1,\cdots,g_N]^T, & \by &= [y_1,\cdots,y_L]^T,\nonumber\\
 \bz &= [\bg^T,\by^T]^T. &  &\label{eq:parm_definitions}
 \end{align}
The number of model parameters to be solved for is now $P = 2(N+L)$, since redundant calibration is a complex problem.
